A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay

### 2020-06-22 Paper/Science/main.tex

\gdef\arcsec{$^{\prime\prime}$}
\def\SNABC{SN Requiem\xspace}

@@

%\title{Requiem for a 10 billion year old supernova } 
\title{A Gravitationally Lensed Supernova with an Observable Decade-long Time Delay}
%\title{A Strongly Lensed Type Ia Supernova with a Decade-Long Time Delay}
%\title{SN Requiem: A Strongly Lensed Type Ia Supernova with a Decade-Long Time Delay}

@@

  We report the discovery of a quadruply-lensed Type Ia supernova (SN) explosion that will produce a time delay measurement uncertainty of $<1\%$. This unprecedented scenario will enable a unique and highly precise measurement of the Hubble constant from a single lensed SN.  The SN is in an evolved galaxy at $z=1.95$, gravitationally lensed by the massive foreground galaxy cluster MACSJ0138. With three lensed images of the supernova detected in existing observations from the \textit{Hubble Space Telescope} with time delays of order 50 days, this is the most distant multiply-imaged supernova yet discovered.  A fourth image in the component closest to the cluster core is predicted to appear in the year 2035$\pm$2.  Observing the light curve of that future image will provide a time delay precision of $\approx 7$ days over an extraordinary baseline of nearly 20 years.

@@

Observations of Type Ia supernova (SN) explosions have played a key role in mapping the cosmic expansion history, and led to the discovery of dark energy that now appears to be driving an accelerating cosmic expansion rate\cite{riess_observational_1998,perlmutter_measurements_1999, Riess_large_2019}.  Determining the nature of dark energy and how it may evolve over time is a primary goal for the large-scale cosmology experiments of the 2020s \cite{amendola_cosmology_2013,spergel_wide_2015,Ivezic_lsst_2019}.
A series of recent investigations into the expansion rate of the universe (the Hubble-LeM\^aitre constant; $H_0$) have found that measurements from the local universe are significantly different from the value inferred from measurement of the cosmic microwave background (CMB) radiation \cite{Riess_large_2019,aghanim_planck_2018}.  Resolving this tension in $H_0$ measurements may reveal new physics from the early universe \textcolor{red}{Citation here?}, and to do so requires multiple 
independent cosmological probes.

@@

Gravitationally lensed SN with multiple images present an attractive new addition to this field. Most notably, SN exhibit relatively simple photometric behavior, with well-understood light curve shapes and colors---a contrast to the stochastic variation of quasars.  However, the exploration of lensed SN time delay cosmography has been hindered by their rarity.  To date, there have been only two lensed SN observed with multiple images \cite{kelly_multiple_2015,goobar_iptf16geu:_2017}. The first, {\it SN Refsdal}, was a peculiar Type II SN whose image with the longest delay was missed \cite{kelly_SNRefsdal_2016}. The second, {\it SN iPTF16geu}, was a Type Ia SN with short delays that make high precision time delay measurements impossible \cite{dhawan_magnification_2020}. Here we present the discovery 
of a third lensed SN resolved into multiple images: SN Requiem.

\todo[inline]{put in precise UT dates for HST data and double check newman proposal \href{https://archive.stsci.edu/hst/abstract.html}{(this one?)}}

We discovered SN Requiem using data from the Hubble Space Telescope (HST) program {\it Resolved Quiescent Magnified Galaxies} (REQUIEM, HST-GO-15663, PI:Akhshik). This project targets rare examples of quiescent galaxies that have been magnified by strong gravitational lensing. %These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular galaxy targeted by REQUIEM
is MRG-M0138, first discovered as a massive red galaxy (MRG) at $z=1.95$ \cite{newman_resolving_2018} behind the massive galaxy cluster MACS J0138.0-2155 \cite{ebeling_macs_2001}.
MRG-M0138 is quadruply lensed by the foreground cluster, which is at $z=0.338$.  
During analysis of REQUIEM observations obtained July 2019 we discovered three point 
sources that were present in archival HST images from July 2016, part of the program 
that originally discovered the MRG-M0138 galaxy (HST-GO-14496; PI:Newman). 
Each point source is within 5 arcseconds of one of the four MRG-M0138 images.  None of the
three point sources are present in the REQUIEM HST data in 2019 (Fig \ref{fig:layout}). We infer that 
the point sources are multiple images of a single astrophysical 
transient in MRG-M0138, most likely a SN. Hereafter we call this multiply-imaged transient {\it SN Requiem}.

To construct a lens model for the MACS J0138.0-2155 cluster we used the publicly available {\tt Lenstool} software \cite{jullo_bayesian_2007, kneib_lenstool_2011} to model the mass distribution in the cluster core as the combination of a cluster scale and multiple galaxy scale potentials.  To avoid unintended bias, we kept the lens model development completely separate from the analysis of the SN.  Only upon independent completion of both were the results combined for the analysis described here.  The input model constraints are the positions and redshifts of the MRG-M0138 galaxy at $z=1.95$ (both the central light peak and the SN location in each image) as well as a multiply-imaged background 
galaxy at $z=0.766$, both having secure spectroscopic redshifts.  
From this model we derive estimates for the lensing magnification and time delay of each of the SN images, including the predicted fourth image (Tab 1).
The lensing model predicts that the SN should appear in the fourth MRG-M0138 image in 20XX+/-X., with a magnification of X. A fifth image will also appear in 20XX+/- but this will be demagnified and likely not observable.

%The range of possible transients for Requiem is broad, encompassing normal SNe, superluminous SNe (galyam luminous 2019), fast transients (berger fast 2013, drout rapidly 2014), and even gravitational microlensing events (rodney two 2018, kelly extreme 2018). 
To realize the cosmological promise of SN Requiem 
we need to estimate the age of each SN image, which in turn constrains the lensing time delays. For this
goal, it is valuable to have a firm determination of the transient’s class. Expected time delays and magnifications from the lens model exclude any of the various rapidly evolving and low-luminosity stellar transient classes, strongly suggesting that it is a SN. The first-order SN distinction remaining is between a Type Ia SN---the explosion of a white dwarf star in a binary system---and a core collapse SN (CCSN)---the end-point of a star with mass $\gtrsim 10 M_{\odot}$. 
The properties of the host galaxy can inform this classification because CCSN are limited to galaxies with young stellar populations. Limits on the specific star formation rate and age for this host, MRG-M0138, show it to be a high-mass but very quiescent and evolved galaxy, %($\log(M_*/M_\odot)=11.7\pm0.2; )$ , 
unlikely to retain any significant population of high-mass stars \cite{newman_resolving_2018}. Based on observed properties of the host galaxy alone, we find a 75\% probability that Requiem is a Type Ia SN (see supplementary text). Adopting the lens model magnifications for the three observed SN images (Table S2) the position of each SN image in color-magnitude space (Fig \ref{fig:class}) provides more classification leverage, yielding p(Ia)=\textcolor{red}{XX}. By also including the measured time delays, we can treat the three SN images effectively as three points on a common SN light curve, and we find p(Ia)=\textcolor{Orange}{YY} (Fig \ref{fig:class}).
%This is corroborated by the limited data available for the transient. In color -magnitude space it displays best agreement with the Type Ia population (Fig 1). 
An improved classification can be achieved with spectroscopy and multi-band photometry upon arrival of the fourth image.

@@

\outline{
Multiple lensed transient or varying sources can be used to constrain cosmology, through timedelay measurements.  Many quasars lensed by galaxies known, provide some of the most accurate measurements of H0.  Possibly in dissagremment with CMB }
\newline
\newline
\outline{
Refsdal proposed multiple lensed SN as means of constraining cosmology. A few have been observed (list), most famously the Refsdal SN, a Type II at z=1.5 lensed by a cluster, missed one image, observed 2 and predicted a fourth that was later observed
}\newline
\newline
\outline{
Requiem is a hubble project that target rare examples of quiescent galaxies, that have been magnified by gravitational lensing. These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular target of REQUIEM, first discovered by Newman in 2016 is RGM0138, is a z=2.X, logM=XX quiescent galaxy quadruply lensed by a cluster at z=0.X. The galaxy has very low SFR=XX and based on stellar ages stopped forming stars ~X.X Gyr earlier. This is a proto typical early type galaxy progenitor.
}
\newline
\newline
\outline{
During analysis of the REQUIEM observations obtained  July 2019 we discovered a new pointsource close to three of the four images, which was not there in the original Hubble images taken in July 2016 (PI Newman), when the system was first discovered (Fig \ref{fig:layout}). 
}
\newline
\newline

\outline{
\todo[inline]{JP: I'll fill this in from the old text, then SR and/or KO work on this classification section} The SN has different colors and brightness in the three observed images. This is expected as SN evolve in brightness and color with time, and the three images represent different times on the supernova light curve, due to the different time delays of the different images. Fig \ref{fig:class} shows the colors and magnitudes expected for different types of SN lightcurves. The observations classify the explosion as a Type Ia, when the magnifications from the lensing model is taken into account. This is consistent with the host galaxy being an evolved early type (void of O and B stars). Using the absolute magnitude, $M_K$, and color, $B-K$, of the host, there is about 75 percent probability that the object is of type Ia (citation to https://arxiv.org/pdf/1309.2630.pdf).
}
\newline
\newline
\outline{
\todo[inline]{JP: Need Johan to fill in this section, or give details for us to fill it in} The lensing model is derived using the lenstool software, constrained by XX spectroscopically confirmed lensed images, and scaling relation for the cluster galaxies, as well as local pertubers .... (johan)
}
\newline
\newline
\outline{
\todo[inline]{JP: I'll fill this in + Johan lens model stuff, this will include time delay/magnification constraints from SNTD} In addition to the relative magnifications of the individual images, the lensing model predicts relative timdelays between the images. These are are XXX in a greement with the observed colors evolution in Fig 2 providing confidence in the accuracy of the lensing model. The lensing model predicts that the SN should appear in the fourth image in 20XX+/-X., with a magnification of X. A fifth image will also appear in 20XX+/- but this will be demagnified and likely not observable.
}
\newline
\newline
\outline{
This is the second highest redshift SNIa, it is the first one that where we observe it from the appearance of the first image to the last (Refsdal had an image that appeared Xyears before it was first dicovered), and it it the one with the longest (predicted) timedelay.  Add here other properties that make the system unique.
}
\newline
\newline
\outline{
\todo[inline]{JP: Can edit time delay/cosmo section} Due to the long timedelay baseline, the sensitivity to cosmology is exceptional, and the precision with which cosmological parameters can be constrained is limited only by the precision of the lensing model. Assuming that H0 is determined from other methods when the fourth image appears in 20XX Fig 3 shows that this single system will have provide comparable (and complementary) constraints as those expectede from tens of lensed supernova in planned LSST and WFIRST cosmological surveys. At that time JWST will likely be in the past but we will have E-ELT, and XXX.. which will allow for cool science such as...
}

The color of a SNIa evolves substantially over its lifetime as the photosphere expands and cools, revealing different layers of the expanding shell and driving episodes of recombination \cite{woosley_type_2007,kasen_type_2009}. Since the phenomenon of gravitational lensing in general is achromatic, this color evolution makes it possible to derive an age constraint that is largely independent of the lens model.\todo{add a lens-model-independent precision statement?}  Combining this information with magnification constraints from the lens modeling helps break parameter degeneracies, yielding the measured delays in Table 1. 
Remarkably, the ages of image 1 and 3 are constrained to better than $\pm$20 days, despite having only a single epoch of photometric data.   These uncertainties may be 
further reduced when the future fourth image is observed with high-precision, multi-epoch photometry.  Such a light curve will pin down the intrinsic SN light curve parameters that are shared by all images.
%These constraints represent an upper limit on the age (and therefore time delay) uncertainties, but they would still deliver an uncertainty of $\sim0.3\%$ on the predicted $\sim6000$ day delay. While already dwarfed by the expected lens model uncertainty, constraints on the intrinsic SN light curve parameters gained from exquisite photometry of the future fourth image of Requiem could improve this even further.

Future large scale surveys such as the Vera C. Rubin Observatory 
%Legacy Survey of Space and Time (LSST) 
and the Nancy Grace Roman Space Telescope will observe dozens to hundreds of lensed SN over their duration \cite{oguri_gravitationally_2010,goldstein_rates_2019,wojtak_magnified_2019}. However, many of these lensed SN will require follow-up resources and all will have significantly shorter delays (and therefore higher relative uncertainties) than SN Requiem. In Figure \ref{fig:cosmo} we imagine a scenario in which scarce follow-up resources could be used to observe $\sim5$ objects. We compare the cosmological constraints obtainable from small samples of SN representative of those expected from the Rubin Observatory and the Roman Space Telescope, with ``Requiem-Like'' SN.

Because of the unusually large ratio of source to lens redshift, the constraints on dark energy equation of state parameters ($w_0$, $w_a$) from the Requiem-like sample are nicely complementary to the samples accessible to the wide-area Rubin Observatory and Roman Space Telescope surveys. The extraordinarily long time delays of cluster-lensed SNe like SN Requiem also deliver significantly better time delay precision, with comparable observational cost.  
Such a sample could be developed with regular monitoring of cluster-scale lenses.  As SN Requiem has shown, even a single epoch of photometric data at the time of detection may be sufficient to provide the necessary anchor point for this promising cosmological tool.

@@

%BibTeX users: After compilation, comment out the following two lines and paste in
% the generated .bbl file. 
\clearpage
\begin{figure*}
    \centering
    \includegraphics[width=0.9\textwidth]{Paper/Figures/fig1_layout.pdf}
    \caption{Overview of the field around the center of the MACSJ0138 cluster field and the locations of the lensed host (H1--4) and \SNABC (SN1--3) images. The wide-field view in $a)$ is 40\arcsec\ on a side with ticks indicating 10\arcsec\ intervals.  Panels b--g) show 4\arcsec\ cutouts around the lensed SN images with 1\arcsec\ ticks.  The three-color images are generated from the WFC3/IR filters as indicated; panels $b$--$d$ show the imaging from July 2016 where the SN was visible and $e$--$g$ show the later imaging from July 2019 where the the SN has faded away.  All panels use the late-epoch F125W imaging for the green channel.  Nevertheless, it is immediately clear that the SN2 image is substantially bluer than the other two, which we use to constrain the transient classification as a likely Ia supernova explosion.}

    \label{fig:layout}
\end{figure*}
\clearpage
\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{Paper/Figures/classification_contours_timeline.pdf}
    \caption{Classification information for \SNABC based on its position in color-magnitude space. (left) Contours show the population distributions for normal SNe of Type Ia (red), Type Ib/Ic (gold), and Type II (green).  These population contours are drawn to enclose 68\% (interior solid lines) and 95\% (exterior shaded regions) of each SN population.  Each SN sub-class was simulated at $z=1.95$, and samples from their expected light curves were drawn uniformly in time. Open markers show the observed photometry of the SN. Dotted vertical lines mark the magnification correction based on the lens modeling, terminating in closed markers with error bars that show the magnitude uncertainty, which is dominated by the lens modeling uncertainty.  (right) The evolution of a Type Ia SN at $z=1.95$ in observed color-mag space is shown by a colored line, with the color indicating SN age in observer-frame days relative to peak brightness.  White diamonds correspond to the times labeled on the colorbar at right. An interactive 3-D version of this figure is available online at \href{https://plot.ly/~jpierel/6/#/}{https://plot.ly/~jpierel/6/#/}
    \label{fig:class}
    }
\end{figure}

\clearpage
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{../Figures/snrequiem_hw_w0wa_apples_to_lsst_ngrst.pdf}
    \caption{ \todo[inline]{replace caption for new figure}The unusual sensitivity of SN Requiem to dark energy equation of state parameters ($w_0, w_a$). Contours show the 1$\sigma$ confidence region for a typical sample of lensed SNIa time delays from LSST (blue) and WFIRST (red) compared to SN Requiem (magenta), which is shown for assumed uncertainties on the lens potential of 5\% and 2\%. Dashed lines and the white marker show the input values of $w_0=-1$, $w_a=0$ that were used for this simulation. In order to isolate the $w_0, w_a$ dependence, all contours are constructed assuming perfect knowledge of $H_0$, $\Omega_{m,0}$ and $\Omega_{\rm de,0}$.  SN Requiem could be highly complementary to the LSST+WFIRST lensed SN sample, due to its unusual combination of a relatively low-redshift lens ($z=0.338$) and high-redshift source ($z=1.95$). This single SN could deliver constraints on dark energy equation of state parameters comparable to several dozen SNe with shorter and less precise  time delays.}
    \label{fig:cosmo}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{crr|rrrr}
    \multicolumn{1}{c}{}&
    \multicolumn{2}{c}{\textbf{Photometry}}&\multicolumn{2}{c}{\textbf{Lens Model}}\\
    %\multicolumn{1}{c}{}&\multicolumn{1}{c}{}\\
    \multicolumn{1}{c}{\textbf{Image}} &\multicolumn{1}{c}{\textbf{Age}} &\multicolumn{1}{c}{$\mathbf{t_i-t_2}$}&\multicolumn{1}{c}{$\mathbf{\mu}$}
    &\multicolumn{1}{c}{$\mathbf{t_i-t_2}$}\\

\midrule

\textit{1}  &$79.10^{+45.56}_{-11.76}$&$91.31^{+62.49}_{-23.36}$ &8.942&103.6\\
\textit{2} & $-16.67^{+19.14}_{-13.10}$&-- \ \ \ \ \ \ \ \ \  & 18.48&-- \ \ \ \ \\
\textit{3} &$80.46^{+20.10}_{-9.84}$&$92.60^{+37.28}_{-21.76}$ & 13.257&153.8\\
\textit{4} &-- \ \ \ \ \ \ \ \ \ &-- \ \ \ \ \ \ \ \ \ &1.137&5980.4\\

\end{tabular}
\caption{\label{tab:time_delays}Time Delays}
\end{table}

@@

\clearpage
\section*{Supplementary materials}
Materials and Methods\\
Supplementary Text\\
Figs. S1 to S10\\
Tables S1 to S2\\
References \textit{(4-10)}
\clearpage
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}

@@

In order to correctly estimate the magnification factors, time delays, and predict the appearance of the future images of MRG0138-SN we need to precisely model the mass distribution in the MRG0138 cluster core. To do so we make use of the latest version of Lenstool \cite{jullo_bayesian_2007}\footnote{publicly available at \url{ https://git-cral.univ-lyon1.fr/lenstool/lenstool}}, which performs a Bayesian analysis with an MCMC sampler to estimate the best fit and error on each parameter of the mass distribution.

>>>>

\gdef\arcsec{$^{\prime\prime}$}
\def\SNABC{SN Requiem\xspace}
\def\reqlike{Requiem-like\xspace}
\def\lenstool{{\tt LENSTOOL}\xspace}

@@

%\title{Requiem for a 10 billion year old supernova } 
\title{A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay}
%\title{A Strongly Lensed Type Ia Supernova with a Decade-Long Time Delay}
%\title{SN Requiem: A Strongly Lensed Type Ia Supernova with a Decade-Long Time Delay}

@@

  We report the discovery of a quadruply-lensed supernova (SN) explosion that will enable a time delay measurement with an uncertainty of $<1\%$. This unprecedented scenario will enable a rare and highly precise measurement of the Hubble constant from a single lensed SN.  The SN is in an evolved galaxy at $z=1.95$, gravitationally lensed by the massive foreground galaxy cluster MACSJ0138.  
 This is the most distant multiply-imaged supernova yet discovered, and we find it is classified as a Type Ia with {$>90\%$} probability.
  Three lensed images of the SN are detected in existing observations from the \textit{Hubble Space Telescope} with relative time delays of $<$200 days.  A fourth image in the component closest to the cluster core is predicted to appear in the year 2037$\pm$2.  Observing the light curve of that future image could provide a time delay precision of $\approx 7$ days over an extraordinary baseline of nearly 20 years.

@@

Observations of Type Ia supernova (SN) explosions have played a key role in mapping the cosmic expansion history, and led to the discovery of dark energy that now appears to be driving an accelerating cosmic expansion rate\cite{riess_observational_1998,perlmutter_measurements_1999, Riess_large_2019}.  Determining the nature of dark energy and how it may evolve over time is a primary goal for the large-scale cosmology experiments of the 2020s \cite{amendola_cosmology_2013,spergel_wide_2015,Ivezic_lsst_2019}.
A series of recent investigations into the expansion rate of the universe (the Hubble-LeM\^aitre constant; $H_0$) have found that measurements from the local universe are significantly different from the value inferred from measurement of the cosmic microwave background (CMB) radiation \cite{Riess_large_2019,aghanim_planck_2018}.  Resolving this tension in $H_0$ measurements may reveal new physics from the early universe, and to do so requires multiple 
independent cosmological probes \cite{verde_tensions_2019}.

@@

Gravitationally lensed SN with multiple images present an attractive new addition to this field. Most notably, SN exhibit relatively simple photometric behavior, with well-understood light curve shapes and colors---a contrast to the stochastic variation of quasars.  However, the exploration of lensed SN time delay cosmography has been hindered by their rarity.  To date, there have been only two lensed SN observed with multiple images \cite{kelly_multiple_2015,goobar_iptf16geu:_2017}. The first, {\it SN Refsdal}, was a peculiar Type II SN whose image with the longest delay was missed \cite{kelly_SNRefsdal_2016}. The second, {\it SN 2016geu}, was a Type Ia SN with short delays that make high precision time delay measurements impossible \cite{dhawan_magnification_2020}. Here we present the discovery 
of a third lensed SN resolved into multiple images: \SNABC.

We discovered SN Requiem using data from the Hubble Space Telescope (HST) program {\it REsolved QUIEscent Magnified Galaxies} (REQUIEM, HST-GO-15663, PI:Akhshik). This project targets rare examples of quiescent galaxies that have been magnified by strong gravitational lensing. %These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular galaxy targeted by REQUIEM
is MRG-M0138, first discovered as a massive red galaxy (MRG) at $z=1.95$ \cite{newman_resolving_2018} behind the massive galaxy cluster MACS J0138.0-2155 \cite{ebeling_macs_2001}.
MRG-M0138 is quadruply lensed by the foreground cluster, which is at $z=0.338$.  
During analysis of REQUIEM observations obtained 13-14 July 2019 \cite{materials_methods_2020} we discovered three point 
sources that were present in archival HST images from 18-19 July 2016, part of the program 
that first confirmed the MRG-M0138 galaxy as a strongly-lensed object (HST-GO-14496; PI:Newman). 
Each point source is within 5 arcseconds of one of the four MRG-M0138 images.  None of the
three point sources are present in the REQUIEM HST data in 2019 (Fig \ref{fig:layout}). We infer that 
the point sources are multiple images of a single astrophysical 
transient in MRG-M0138, most likely a SN. Hereafter we call this multiply-imaged transient \SNABC.

To construct a lens model for the MACS J0138.0-2155 cluster we used the publicly available \lenstool software \cite{jullo_bayesian_2007, kneib_lenstool_2011} to model the mass distribution in the cluster core as the combination of a cluster scale and multiple galaxy scale potentials \cite{materials_methods_2020}.  To avoid unintended bias, we kept the lens model development completely separate from the analysis of the SN.  Only upon independent completion of both were the results combined for the analysis described here.  The input model constraints are the positions and redshifts of the MRG-M0138 galaxy at $z=1.95$ (both the central light peak and the SN location in each image) as well as a multiply-imaged background 
galaxy at $z=0.766$, both having secure spectroscopic redshifts \cite{materials_methods_2020}.  
From this model we derive estimates for the lensing magnification and time delay of each of the SN images, including the predicted fourth image (Table \ref{tab:time_delays}).
The lensing model predicts that the SN should appear in the fourth MRG-M0138 image in the year 2037$\pm$1.5, demagnified with $\mu=0.4\pm0.2$. A fifth image will also appear at a later date, located very near the center of the cluster and much more significantly demagnified, so therefore likely not observable.

%The range of possible transients for Requiem is broad, encompassing normal SNe, superluminous SNe (galyam luminous 2019), fast transients (berger fast 2013, drout rapidly 2014), and even gravitational microlensing events (rodney two 2018, kelly extreme 2018). 
To realize the cosmological promise of SN Requiem 
we need to estimate the age of each SN image, which in turn constrains the lensing time delays. For this
goal, it is valuable to have a firm determination of the transient’s class. Expected time delays and magnifications from the lens model exclude any of the various rapidly evolving and low-luminosity stellar transient classes, strongly suggesting that it is a SN. The first-order SN distinction remaining is between a Type Ia SN---the explosion of a white dwarf star in a binary system---and a core collapse SN (CCSN)---the end-point of a star with mass $\gtrsim 10 M_{\odot}$. 
The properties of the host galaxy can inform this classification because CCSN are limited to galaxies with young stellar populations. Limits on the specific star formation rate and age for this host, MRG-M0138, show it to be a massive but very quiescent and evolved galaxy, %($\log(M_*/M_\odot)=11.7\pm0.2; )$ , 
unlikely to retain any significant population of high-mass stars \cite{newman_resolving_2018}. Based on observed properties of the host galaxy alone, we find a 62-75\% probability that \SNABC is a Type Ia SN \cite{materials_methods_2020}. Adopting the lens model magnifications for the three observed SN images (Table \ref{tab:time_delays}) the position of each SN image in color-magnitude space (Figure \ref{fig:class}) provides more classification leverage, yielding p(Ia)=95\% \cite{materials_methods_2020}. By also including the model-predicted time delays, we can treat the three SN images effectively as three points on a common SN light curve, and we find p(Ia)=94\% \cite{materials_methods_2020}.
%This is corroborated by the limited data available for the transient. In color -magnitude space it displays best agreement with the Type Ia population (Fig 1). 
An improved classification can be achieved with spectroscopy and multi-band photometry upon arrival of the fourth image.

@@

The color of a SNIa evolves substantially over its lifetime as the photosphere expands and cools, revealing different layers of the expanding shell and driving episodes of recombination \cite{woosley_type_2007,kasen_type_2009}. Since the phenomenon of gravitational lensing in general is achromatic, this color evolution makes it possible to derive an age constraint that is largely independent of the lens model \cite{materials_methods_2020}.  Combining this information with magnification constraints from the lens modeling helps break parameter degeneracies, yielding the measured delays in Table \ref{tab:time_delays} \cite{materials_methods_2020}. 
Remarkably, the ages of image 1 and 3 are constrained to better than $\pm$20 days, despite having only a single epoch of photometric data.    These uncertainties may be 
further reduced when the future fourth image is observed with high-precision, multi-epoch photometry.  Such a light curve will pin down the intrinsic SN light curve parameters that are shared by all images, and break remaining parameter degeneracies (see supplemental materials).
%These constraints represent an upper limit on the age (and therefore time delay) uncertainties, but they would still deliver an uncertainty of $\sim0.3\%$ on the predicted $\sim6000$ day delay. While already dwarfed by the expected lens model uncertainty, constraints on the intrinsic SN light curve parameters gained from exquisite photometry of the future fourth image of Requiem could improve this even further.

Future large-scale surveys such as the Vera C. Rubin Observatory 
%Legacy Survey of Space and Time (LSST) 
and the Nancy Grace Roman Space Telescope will observe dozens to hundreds of lensed SN over their duration \cite{oguri_gravitationally_2010,goldstein_rates_2019,wojtak_magnified_2019}. However, many of these lensed SN will require substantial follow-up resources to be cosmologically useful. Furthermore, almost all will have significantly shorter 
delays (of order 10-100 days) and therefore higher relative uncertainties than SN Requiem.  Since it is the {\it fractional} time delay uncertainty that 
propagates through to any time delay distance measurement, the extraordinarily long time delays of cluster-lensed SNe like SN Requiem can deliver significantly better time delay precision, with comparable observational cost.  In fact, the long time baseline of lensed SN like \SNABC effectively insures that their cosmological precision is not limited by the time delay measurement errors.

In Figure \ref{fig:cosmo} we illustrate the cosmological value of time delay cosmography with objects like \SNABC  by considering a small sample of 5 ``\reqlike'' SN.  For a comparison on equal footing, we evaluate equal-sized lensed SN samples 
that mimic those expected from the Rubin Observatory and  the Roman Space Telescope.  Differences in the 
ratio of source to lens redshift account for the different cosmological parameter covariances, reflected in the orientations of the contours in these plots. Because of the unusually large redshift ratio of SN Requiem, the constraints 
%on dark energy equation of state parameters ($w_0$, $w_a$) 
from the \reqlike sample are nicely complementary to the low-$z$ Rubin Observatory and high-$z$ Roman Space Telescope samples. 
Furthermore, when we assume that the lens modeling uncertainty is equal for each sample (amounting to a 5\% uncertainty in each time delay distance), the \reqlike sample has leverage that is comparable to the same number of lensed SN from the wide-field surveys. However, if we improve  the lensing constraints to 2\% precision, the \reqlike contours tighten dramatically, reflecting the fact that these long-time-baseline events are limited only by the lens modeling uncertainty.

To achieve these cosmological constraints in practice, such a sample could be developed with regular monitoring of cluster-scale lenses, partnered with modest follow-up to characterize any lensed SN discovered.  As \SNABC has shown, even a single epoch of photometric data at the time of detection may be sufficient to provide the necessary anchor point for this promising cosmological tool.  After that, all we need is patience.

@@

%BibTeX users: After compilation, comment out the following two lines and paste in
% the generated .bbl file. 
\clearpage
\begin{figure*}
    \centering
    \includegraphics[draft=False,width=0.9\textwidth]{Paper/Figures/fig1_layout.pdf}
    \caption{Overview of the MACSJ0138 cluster field and the locations of the SN Requiem (SN1--3) images and its multiply-imaged host galaxy (H1--4). The wide-field view in $a)$ is 40\arcsec\ on a side with ticks indicating 10\arcsec\ intervals.  Panels b--g) show 4\arcsec\ cutouts around the lensed SN images with 1\arcsec\ ticks.  The three-color images are generated from the WFC3/IR filters as indicated; panels $b$--$d$ show the imaging from July 2016 where the SN was visible and $e$--$g$ show the later imaging from July 2019 where the the SN has faded away.  All panels use the late-epoch F125W imaging for the green channel.  Nevertheless, it is immediately clear that the SN2 image is substantially bluer than the other two, which we use to constrain the transient classification as a likely Ia supernova explosion and to constrain the relative age of each SN image.}

    \label{fig:layout}
\end{figure*}
\clearpage
\begin{figure}[h]
    \centering
    \includegraphics[draft=False, width=\textwidth]{Paper/Figures/classification_contours_timeline.pdf}
    \caption{Classification information for SN Requiem based on its position in color-magnitude space. (left) Contours show the population distributions for normal SNe of Type Ia (red), Type Ib/Ic (gold), and Type II (green).  These population contours are drawn to enclose 68\% (interior solid lines) and 95\% (exterior shaded regions) of each SN population.  Each SN sub-class was simulated at $z=1.95$, and samples from their expected light curves were drawn uniformly in time. Open markers show the observed photometry of the SN. Dotted vertical lines mark the magnification correction based on the lens modeling, terminating in closed markers with error bars that show the magnitude uncertainty, which is dominated by the lens modeling uncertainty.  (right) The evolution of a typical Type Ia SN at $z=1.95$ in observed color-mag space is shown by a colored line, with the color indicating SN age in observer-frame days relative to peak brightness.  White diamonds correspond to the times labeled on the colorbar at right. Grey shading shows the typical range of luminosities observed for the Type Ia SN population in the nearby universe \cite{wang_determination_2006}. An interactive 3-D version of this figure is available online at \href{https://plot.ly/~jpierel/6/#/}{https://plot.ly/~jpierel/6/#/}
    \label{fig:class}
    }
\end{figure}

\clearpage
\begin{figure}
    \centering
    \includegraphics[draft=False,trim={2cm, 3cm, 3cm, 4cm},clip,width=\textwidth]{Paper/Figures/snrequiem_hw_w0wa_apples_to_lsst_ngrst_4panel.pdf}
    \caption{  
    The sensitivity of SN Requiem to the expansion rate of the universe ($H_0$) and dark energy equation of state parameters ($w_0, w_a$). Dashed lines and the white marker show the input values (upper right of the top panels) that were used for this simulation. Contours show the 1$\sigma$ confidence region for a projected sample of 5 lensed SNIa time delays from the Rubin Observatory (blue) and the Roman Space Telescope (red) compared to 5 events like \SNABC (black).  The left and right columns assume a 5\% and 2\% contribution to the distance uncertainty from modeling the lens potential, respectively. Improving the lens modeling tightens the cosmological constraints for all three samples, but the improvement for the \SNABC-like sample is the most significant.  This is simply because the long time delay baseline for \SNABC makes the time delay uncertainty negligible, raising the ceiling for cosmological improvement with tighter lens modeling constraints.  In order to isolate the plotted parameters, all contours are constructed assuming perfect knowledge of the cosmological parameters that are not shown. 
    }
    \label{fig:cosmo}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{ccc|cc}
    \multicolumn{1}{c}{}&
    \multicolumn{2}{c}{\textbf{Inferred from Photometry}}&\multicolumn{2}{c}{\textbf{Lens Model Predictions}}\\
    %\multicolumn{1}{c}{}&\multicolumn{1}{c}{}\\
    \multicolumn{1}{c}{\textbf{Image}} &\multicolumn{1}{c}{\textbf{Age}} &\multicolumn{1}{c}{$\mathbf{t_i-t_1}$}&\multicolumn{1}{c}{$\mathbf{\mu}$}
    &\multicolumn{1}{c}{$\mathbf{t_i-t_1}$}\\

\midrule
\textit{1}  & $112^{+20}_{-38}$ & -- & $3.9\pm0.5$ & --\\
\textit{2} & $-27^{+26}_{-6}$ & $139^{+63}_{-23}$ & $7\pm3$ & 101\pm62 \\
\textit{3} & $88\pm15$ & $16\pm19$ & $5\pm1$ & 19\pm34 \\
\textit{4} & -- & -- & $0.4\pm0.2$ & 7761\pm541\\

% 2020.06.19  OLD DATA (prior to MODEL H) with time delays referenced to image 2.
%\textit{1}  &$79.10^{+45.56}_{-11.76}$&$91.31^{+62.49}_{-23.36}$ &8.942&103.6\\
%\textit{2} & $-16.67^{+19.14}_{-13.10}$&-- \ \ \ \ \ \ \ \ \  & 18.48&-- \ \ \ \ \\
%\textit{3} &$80.46^{+20.10}_{-9.84}$&$92.60^{+37.28}_{-21.76}$ & 13.257&153.8\\
%\textit{4} &-- \ \ \ \ \ \ \ \ \ &-- \ \ \ \ \ \ \ \ \ &1.137&5980.4\\
\end{tabular}
\caption{\label{tab:time_delays}Time Delays.  The combination of the models for the light curve evolution and the lens time delays implies a date of $2037\pm2$ for the peak brightness of Image 4.}
\end{table}

@@

\clearpage
\section*{Supplementary materials}
Materials and Methods\\
Supplementary Text\\
Figs. S1 to S9\\
Tables S1 to S7\\
References \textit{(4-10)}
\clearpage
\setcounter{table}{0}
\renewcommand{\thetable}{S\arabic{table}}

@@

In order to correctly estimate the magnification factors, time delays, and predict the appearance of the future images of MRG0138-SN we need to precisely model the mass distribution in the MRG0138 cluster core. To do so we make use of the latest version of \lenstool \cite{jullo_bayesian_2007}\footnote{publicly available at \url{ https://git-cral.univ-lyon1.fr/lenstool/lenstool}}, which performs a Bayesian analysis with an MCMC sampler to estimate the best fit and error on each parameter of the mass distribution.
